---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
if (!requireNamespace("dplyr", quietly = TRUE)) {
  warning(call. = FALSE, "This vignette requires dplyr.")
  knitr::knit_exit()
}

if (!requireNamespace("ivs", quietly = TRUE)) {
  warning(call. = FALSE, "This vignette requires dplyr.")
  knitr::knit_exit()
}

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Motivation

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/TimTaylor/episodes/workflows/R-CMD-check/badge.svg)](https://github.com/TimTaylor/episodes/actions)
<!-- badges: end -->

***episodes*** is a (hopefully) temporary solution to a small inconvenience
that relates to [data.table](https://cran.r-project.org/package=data.table),
[dplyr](https://cran.r-project.org/package=dplyr) and
[ivs](https://cran.r-project.org/package=ivs); namely that dplyr is currently
[slow when working with a large number of groupings](https://github.com/tidyverse/dplyr/issues/5017)
and data.table [does not support the record class](https://github.com/Rdatatable/data.table/issues/4910)
on which ivs intervals are based.

To expand on issues consider the following small set of episode data:

```{r, message=FALSE}
library(episodes)
library(dplyr)
library(ivs)
library(data.table)

id1 = c(1,1,2,2,2,1)
start1 = as.Date(c("2020-01-01", "2020-01-03","2020-04-01", "2020-04-15", "2020-04-17", "2020-05-01"))
end1 = as.Date(c("2020-01-10", "2020-01-10", "2020-04-30", "2020-04-16", "2020-04-19", "2020-10-01"))
(dat <- tibble(id=id1,start=start1,end=end1))
```

ivs provides an elegant way to find the minimum spanning interval across
these episodes:

```{r}
dat |>
    mutate(interval = iv(start, end + 1)) |>
    group_by(id) |>
    summarise(interval=iv_groups(interval, abutting = FALSE), .groups = "drop")
```

This is great when we only have a small number of ids to group by but is
noticeably slow for a larger number:

```{r}
n=125000
id2 <- sample(seq_len(n), size = n * 5, replace = TRUE)
start2 <- as.Date("2020-01-01") + sample.int(365, size = n*5, replace = TRUE)
end2 <- start2 + sample(1:100, size = n*5, replace = TRUE)
(big_dat <- tibble(id=id2, start=start2,end=end2))

system.time(
    big_dat |>
        mutate(interval = iv(start, end + 1)) |>
        group_by(id) |>
        summarise(interval=iv_groups(interval, abutting = FALSE), .groups = "drop") ->
        out
        
)
```

If you were not already using it, this is likely the time you would reach for
the data.table package. Unfortunately the interval class created by ivs is built
upon on the [record type from vctrs](https://vctrs.r-lib.org/reference/new_rcrd.html),
and this class is not supported in data.table:

```{r, error=TRUE}
    DT <- as.data.table(big_dat)
    DT[, interval:=iv(start, end+1)]
```

***episodes*** solves this with the `merge_episodes()` function:

```{r}
system.time(
    big_dat |> 
        merge_episodes() |> 
        mutate(interval = iv(start = .episode_start, end = .episode_end + 1)) ->
        out2
)

# check for equality
all.equal(out, select(out2, id, interval))
```

We also provide another function `add_parent_interval()` that associates the
the minimum spanning interval with each observation without reducing to the
unique values:

```{r}
add_parent_interval(big_dat)
```
