---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
if (!requireNamespace("dplyr", quietly = TRUE)) {
    warning(call. = FALSE, "This README requires dplyr >= v1.1.0.")
    knitr::knit_exit()
}

if (!requireNamespace("ivs", quietly = TRUE)) {
    warning(call. = FALSE, "This README requires ivs.")
    knitr::knit_exit()
}

knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    fig.path = "man/figures/README-",
    out.width = "100%"
)
```

# NHSRepisodes

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![R-CMD-check](https://github.com/nhs-r-community/NHSRepisodes/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/nhs-r-community/NHSRepisodes/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Installation instructions

You can install the development version of this package from GitHub with:

``` r
# install.packages("remotes")
remotes::install_github("https://github.com/nhs-r-community/NHSRepisodes")
```

## Motivation

***NHSRepisodes*** is a (hopefully) temporary solution to a small inconvenience
that relates to [data.table](https://cran.r-project.org/package=data.table),
[dplyr](https://cran.r-project.org/package=dplyr) and
[ivs](https://cran.r-project.org/package=ivs); namely that dplyr is currently
[slow when working with a large number of groupings](https://github.com/tidyverse/dplyr/issues/5017)
and data.table [does not easily support the record class](https://github.com/Rdatatable/data.table/issues/4910)
on which ivs intervals are based.

To expand on issues consider the following small set of episode data:

```{r, message=FALSE}
library(NHSRepisodes)
library(dplyr)
library(ivs)
library(data.table)

# note - we need functionality introduced in dplyr 1.1.0.
if (getNamespaceVersion("dplyr") < "1.1.0") {
    warning("Please update dplyr to version 1.1.0 or higher to run these examples.")
    knitr::knit_exit()
}

# Let's note the package versions used in generating this README
packages <- c("NHSRepisodes", "dplyr", "data.table", "ivs")
mutate(tibble(packages), version = sapply(packages, getNamespaceVersion))

# Create a dummy data set give the first and last dates of an episode
dat <- tribble(
    ~id, ~start, ~end,
    1L, "2020-01-01", "2020-01-10",
    1L, "2020-01-03", "2020-01-10",
    2L, "2020-04-01", "2020-04-30",
    2L, "2020-04-15", "2020-04-16",
    2L, "2020-04-17", "2020-04-19",
    1L, "2020-05-01", "2020-10-01",
    1L, "2020-01-01", "2020-01-10",
    1L, "2020-01-11", "2020-01-12",
)

# This will create an object called dat and also open in the console
(dat <- mutate(dat, across(start:end, as.Date)))
```

The {ivs} package provides an elegant way to find the minimum spanning interval 
across these episodes:

```{r}
dat |>
    mutate(interval = iv(start = start, end = end + 1)) |>
    reframe(interval = iv_groups(interval, abutting = FALSE), .by = id)
```

Note that {ivs} creates intervals that are *right-open* meaning they are
inclusive on the left (have an opening square bracket `[`) and exclusive on the
right (with a closing a rounded bracket `)`). Consequently, in our first call to
`mutate()` we added 1 to the `end` value. This ensures that the full range of
dates are considered (e.g. for the first row we want to consider all days from
`2020-01-01` to `2020-01-10` not only up until `2020-01-09`).

This works great when we only have a small number of ids to group by. However,
it becomes noticeably slow for a larger number:

```{r}
# Creating a larger data set
n <- 125000
id2 <- sample(seq_len(n), size = n * 5, replace = TRUE)
start2 <- as.Date("2020-01-01") + sample.int(365, size = n * 5, replace = TRUE)
end2 <- start2 + sample(1:100, size = n * 5, replace = TRUE)

# creates the object big_dat and shows the first 10 rows as a tibble in the console
(big_dat <- tibble(id = id2, start = start2, end = end2))

# checking the time to run
system.time(
    out_dplyr <- 
        big_dat |>
        mutate(interval = iv(start, end + 1)) |>
        reframe(interval = iv_groups(interval, abutting = FALSE), .by = id)
)
```

If you were not already using it, this is likely the time you would reach for
the {data.table} package. Unfortunately the interval class created by {ivs} is 
built upon on the
[record type from vctrs](https://vctrs.r-lib.org/reference/new_rcrd.html),
and this class is not supported in {data.table}:

```{r, error=TRUE}
DT <- as.data.table(big_dat)
DT[, interval := iv(start, end + 1)]
```

We can go through a few more steps to get a comparable answer but still find
slightly slower performance:

```{r}
fun <- function(s, e) {
    interval <- iv(s, e)
    groups <- iv_groups(interval, abutting = FALSE)
    list(start = iv_start(groups), end = iv_end(groups))
}

system.time(out_dt <- DT[, fun(start, end + 1), by = id])
```

***NHSRepisodes*** solves this with the `merge_episodes()` function:

```{r}
merge_episodes(big_dat)

# And for comparison with earlier timings
system.time(out <- merge_episodes(big_dat))

# equal output (subject to ordering)
out <- out |> 
    mutate(interval = iv(start = .episode_start, end = .episode_end + 1)) |> 
    select(id, interval)

out_dplyr <- arrange(out_dplyr, id, interval)

out_dt <- out_dt |> 
    as.data.frame() |> 
    as_tibble() |> 
    mutate(interval = iv(start = start, end = end)) |> 
    select(id, interval) |> 
    arrange(id, interval)

all.equal(out, out_dplyr)
all.equal(out, out_dt)
```

We also provide another function `add_parent_interval()` that associates the
the minimum spanning interval with each observation without reducing to the
unique values:

```{r}
add_parent_interval(dat)
```
